Iterative Monomorphisation

### 2025-05-15 iterative_monomorphisation.tex
minor

Automatic theorem provers provide automation for proof assistant users. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae. By contrast, many automatic provers operate only on monomorphic logics. One approach to closing this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}. However, this entails a lot of work that must be redone for every prover.

@@

Any translation method based on instantiation will inevitably be incomplete.
By a type version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of \(\varphi\). However, this set cannot be computed~\cite[Theorem~1]{expr-poly-types}.

B\"ohme's iterative approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}. This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers. However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
\texttt{SMT-LIB} makes use of this monomorphisation approach \cite{smt-lib-mono}.
A similar algorithm also appears to be implemented in the \texttt{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.

>>>>

Automatic theorem provers provide automation for proof assistant users. Many proof assistants, such as HOL4~\cite{slind-norrish-2008}, HOL Light~\cite{harrison-2009} and Isabelle\slash HOL \cite{nipkow-et-al-2002}, support rank-1 polymorphism, where type quantification is allowed at the top level of formulae. By contrast, many automatic provers operate only on monomorphic logics. One way to closing this gap is to extend provers to natively support polymorphism, as has been done in Vampire~\cite{bhayat-reger-2020}. However, this entails a lot of work that must be redone for every prover.

@@

Any translation method based on instantiation will inevitably be incomplete.
By a type version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of~\(\varphi\). However, this set cannot be computed in general~\cite[Theorem~1]{expr-poly-types}.

B\"ohme's iterative approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}. This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers. However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
\relax{SMT-LIB} makes use of this monomorphisation approach \cite{smt-lib-mono}.
A similar algorithm also appears to be implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
